\bf Comparing the Power of Plot Designs to Reveal Correlation

### 2015-12-10 paper/paper.Rnw
Figures

Figure 7 shows the raw results of the survey and Figure 8 shows the difference in power with a $95\%$ confidence interval calculated according to the formula in \ref{eq:2}. Values above zero indicate that the scatter plot had more power to detect the true plot than the overlaid lines.
Performance between the two plot designs appears to be similar for positive correlation, with the scatter plot power gradually improving relative to the overlaid lines as sample size increases. The scatter plot appears to be the superior choice as it retains its power with negative correlation, where the overlaid lines performance suffered. This is not surprising, as positive and negative correlation appear as symmetric graphs in the scatter plot but the two characteristics appear very differently in the overlaid line graph (Figure 6). The right side shows the performance benefits of smoothing overlaid line time series data. Much of the 'noise' in the time series is removed and studying the overall trend shows a greatly improved ability to successfully detect the true data plot when it is negatively correlated. The scatter plot still appears to be the superior plot, but much of the gap between the two is reduced for positive correlation.

@@

Figure 9 demonstrates the effects of changing sample size for unsmoothed data. The scatter plot has increased power to detect correlation almost universally for any increase. Sample size influences the overlaid lines differently, they appears to benefit for positively correlated true plots, but it does not improve the weak performance in negative correlation.
Figures 10 and 11 give a breakdown of the interactions between the response variables: plot detection, time and confidence. Figure 8 is the histogram of the time taken to record a response, split by plot design and sign of correlation and true plot detection. The time data was highly positively skewed with a maximum of 1760 seconds so the variable was log-transformed. We found that the subject could identify the true plot out of the overlaid line graph 33.6\% of the time and it took on average $e^{3.442}$ = 31.2 seconds to make a choice. Successful detections were slightly faster than unsuccessful decision, taking on average 28.7 seconds to pick the true plot and 32.6 seconds if a null was selected. The scatter plots performed much better, with the true plot picked 52.9\% of the time, with an average of 22.7 seconds required to make a decision. If the true plot was detected it took only 18.6 seconds, and 29.3 seconds for a null plot selection.

@@

Next we analysed the individual ability of each of our subjects. Figure 12 plots the accuracy of each subject on the y axis versus the average confidence on the x axis. While we expected that subjects with greater visual skills would report lower values of confidence, there does not appear to be any relationship between the two. The accuracy is normally distributed (Jarque-Bera p value = 0.418), with a mean of 42.5\% and a standard deviation of 17.8\%. The strongest subject successfully identified the true plot 91\% of the time, whilst the weakest was never able to identify the true plot. As all subjects were shown a balanced mix of true plot correlations, the extreme difference in individual visual ability is striking. Visual ability will be controlled via a subject specific random effect in the models in the next section.

@@

There are two major reasons why a subject may fail to detect the true plot out of a lineup. Either the plot design was unable to display correlation for that particular combination of factors (and the analyst would commit a Type II error by not rejecting the null hypothesis), or a null plot that had a greater correlation than the true plot was selected instead (and the analyst would commit a Type I error by rejecting the null hypothesis). We next can analyse these kinds of errors with the pin plots found in Figure 13. The frequency of selection of each plot in each lineup, either true or null, is represented by a pin, placed at the value of the correlation between the two variables shown in that plot. The correlation of the selected plot is on the x-axis and the frequency of selection is plotted on the y axis. The correlation of the simulated (true) data (r=-0.9, -0.7, -0.3, 0.3, 0.7, 0.9) in the plot, the sample size (t=12, 24, 48, 96), and the plot design (unsmoothed/smoothed, scatterplot/lineplot) is used to facet the plots. The blue pins indicate the true plots and the red pins indicate the null plots. If the plot design has the power to display correlation, we would expect the highest correlation plot to be selected, leading to a Type I error where a plot from the null distribution spuriously had high correlation. We find that this is generally not the case. The selections appear to have a wide range of correlations, indicating that if a subject selected a null plot it was typically not due to excess spurious correlation. Instead the failure to select the highest correlated plot indicates that a Type II error was committed rather than Type I as the plot designs did not have the power to differentiate between medium and low correlation plots. However, for the scatter plot, whenever there was a plot with a very strong correlation, usually in the form of the true plot, it was overwhelmingly selected. For the overlaid lines this mostly occurred only for plots with strong positive correlation, which is consistent with its general lack of power to reveal negative relationships in data.

@@

Figures 14 and 15 show lineups where a null plot with a strong pattern led to its selection over the data plot.  Mostly, when a null plot was chosen over the data plot it was because the correlation was stronger, but in a few lineups a null plot with small correlation was selected by many subjects.  The true data plot is in position 4, but 9/11 subjects chose plot 6. Plot 6 is unusual, most of the points are scattered around the vertical median, but there are two outlying points, one with a high y-value and one with a low y-value.  Other concurrent work has suggested that outliers are pre-attentive patterns and this plot selection by the observers is consistent with that. Figure 15 below is another example of where subjects chose a null plot that had smaller correlation than the data plot. Four out of five evaluations selected the null plot in position 13,  instead of the data plot in position 8. Plot 13 has a correlation of 0.225, smaller (in absolute value) than the -0.707 of plot 8.  This selection is in-line with Wertheimer’s gestalt law of common fate. It appears to be strongly correlated in the beginning of the series but breaks apart towards the end. The selection implies that subjects selecting this plot focused on the joint path of the two series early in the plot and ignored the contradictory evidence of no relationship as they diverged later in the plot.  It also confounded by the seeming preference to find positive correlation over negative correlation, and we suspect that if one of the series in plot 8 had the values flipped, then the strong positive correlation would have been noticed.

@@

Table 1 summarises the model fits. It can be confusing the interpret the model estimates, and at first glance it appears the overlaid lines (-0.660) are substantially more powerful than the scatter plot (-4.403), but this is misleading, and  interpreting the the full model involves combining the estimates for the different interactions with correlation and sign. Taking correlation into account, the scatter plot shows superior performance. It needs to be noted that we only tested correlations at  0.3, 0.5, 0.7 and 0.9, so the results below an absolute value of 0.3 are unreliable and may be artifacts of regression smoothing methods. The correlation:sign interaction shows that the scatter plot has close to symmetric performance for positive (correlation:sign + correlation) and negative (-correlation) correlation while the overlaid line graph is substantially weaker for negative values of correlation. Surprisingly, smoothing out the excess 'noise' from a time series had a negative effect on scatter plots, despite a minimal visual impact of smoothed (Figure 5). It did have a significant positive impact on the power of overlaid line graphs. Increasing the sample size, $t$, improved the power of the scatter plot but it was not statistically significant for the overlaid line graph.
Figure 16 shows the fitted unrestricted model, with power predictions (y axis) plotted against correlation (x axis) for each of the four sample sizes tested. The thicker lines represent the fixed effect fit, or the average performance across all tested individuals whilst the thinner lines represent the random effects fit, the variation in power that can be attributed to an individual's visual skill. Previous studies [15] show that this random visual skill is not impacted by demographics such as age and gender; whilst education increases the power of a graphic display by only approximately one percent, not a practically significant amount. We did not include these characteristics of the individual in the modelling due to this lack of significance. Figure 16 especially highlights the weak performance of the overlaid lines and the benefits of smoothing.

>>>>

Figure~\ref{results} shows the raw results of the survey and Figure~\ref{powerdiff} shows the difference in power with a $95\%$ confidence interval calculated according to the formula in \ref{eq:2}. Values above zero indicate that the scatter plot had more power to detect the true plot than the overlaid lines.
Performance between the two plot designs appears to be similar for positive correlation, with the scatter plot power gradually improving relative to the overlaid lines as sample size increases. The scatter plot appears to be the superior choice as it retains its power with negative correlation, where the overlaid lines performance suffered. This is not surprising, as positive and negative correlation appear as symmetric graphs in the scatter plot but the two characteristics appear very differently in the overlaid line graph (Figure~\ref{designs}). The right side shows the performance benefits of smoothing overlaid line time series data. Much of the 'noise' in the time series is removed and studying the overall trend shows a greatly improved ability to successfully detect the true data plot when it is negatively correlated. The scatter plot still appears to be the superior plot, but much of the gap between the two is reduced for positive correlation.

@@

Figure~\ref{samplesize} demonstrates the effects of changing sample size for unsmoothed data. The scatter plot has increased power to detect correlation almost universally for any increase. Sample size influences the overlaid lines differently, they appears to benefit for positively correlated true plots, but it does not improve the weak performance in negative correlation.
Figure~\ref{time} and Figure~\ref{confidence} give a breakdown of the interactions between the response variables: plot detection, time and confidence. Figure 8 is the histogram of the time taken to record a response, split by plot design and sign of correlation and true plot detection. The time data was highly positively skewed with a maximum of 1760 seconds so the variable was log-transformed. We found that the subject could identify the true plot out of the overlaid line graph 33.6\% of the time and it took on average $e^{3.442}$ = 31.2 seconds to make a choice. Successful detections were slightly faster than unsuccessful decision, taking on average 28.7 seconds to pick the true plot and 32.6 seconds if a null was selected. The scatter plots performed much better, with the true plot picked 52.9\% of the time, with an average of 22.7 seconds required to make a decision. If the true plot was detected it took only 18.6 seconds, and 29.3 seconds for a null plot selection.

@@

Next we analysed the individual ability of each of our subjects. Figure~\ref{confidence2} plots the accuracy of each subject on the y axis versus the average confidence on the x axis. While we expected that subjects with greater visual skills would report lower values of confidence, there does not appear to be any relationship between the two. The accuracy is normally distributed (Jarque-Bera p value = 0.418), with a mean of 42.5\% and a standard deviation of 17.8\%. The strongest subject successfully identified the true plot 91\% of the time, whilst the weakest was never able to identify the true plot. As all subjects were shown a balanced mix of true plot correlations, the extreme difference in individual visual ability is striking. Visual ability will be controlled via a subject specific random effect in the models in the next section.

@@

There are two major reasons why a subject may fail to detect the true plot out of a lineup. Either the plot design was unable to display correlation for that particular combination of factors (and the analyst would commit a Type II error by not rejecting the null hypothesis), or a null plot that had a greater correlation than the true plot was selected instead (and the analyst would commit a Type I error by rejecting the null hypothesis). We next can analyse these kinds of errors with the pin plots found in Figure~\ref{pins}. The frequency of selection of each plot in each lineup, either true or null, is represented by a pin, placed at the value of the correlation between the two variables shown in that plot. The correlation of the selected plot is on the x-axis and the frequency of selection is plotted on the y axis. The correlation of the simulated (true) data (r=-0.9, -0.7, -0.3, 0.3, 0.7, 0.9) in the plot, the sample size (t=12, 24, 48, 96), and the plot design (unsmoothed/smoothed, scatterplot/lineplot) is used to facet the plots. The blue pins indicate the true plots and the red pins indicate the null plots. If the plot design has the power to display correlation, we would expect the highest correlation plot to be selected, leading to a Type I error where a plot from the null distribution spuriously had high correlation. We find that this is generally not the case. The selections appear to have a wide range of correlations, indicating that if a subject selected a null plot it was typically not due to excess spurious correlation. Instead the failure to select the highest correlated plot indicates that a Type II error was committed rather than Type I as the plot designs did not have the power to differentiate between medium and low correlation plots. However, for the scatter plot, whenever there was a plot with a very strong correlation, usually in the form of the true plot, it was overwhelmingly selected. For the overlaid lines this mostly occurred only for plots with strong positive correlation, which is consistent with its general lack of power to reveal negative relationships in data.

@@

Figures~\ref{anom1} and \ref{anom2} show lineups where a null plot with a strong pattern led to its selection over the data plot.  Mostly, when a null plot was chosen over the data plot it was because the correlation was stronger, but in a few lineups a null plot with small correlation was selected by many subjects. In Figure~\ref{anom1}, the true data plot is in position 4, but 9/11 subjects chose plot 6. Plot 6 is unusual, most of the points are scattered around the vertical median, but there are two outlying points, one with a high y-value and one with a low y-value.  Other concurrent work has suggested that outliers are pre-attentive patterns and this plot selection by the observers is consistent with that. Figure~\ref{anom2} above is another example of where subjects chose a null plot that had smaller correlation than the data plot. Four out of five evaluations selected the null plot in position 13,  instead of the data plot in position 8. Plot 13 has a correlation of 0.225, smaller (in absolute value) than the -0.707 of plot 8.  This selection is in-line with Wertheimer’s gestalt law of common fate. It appears to be strongly correlated in the beginning of the series but breaks apart towards the end. The selection implies that subjects selecting this plot focused on the joint path of the two series early in the plot and ignored the contradictory evidence of no relationship as they diverged later in the plot.  It also confounded by the seeming preference to find positive correlation over negative correlation, and we suspect that if one of the series in plot 8 had the values flipped, then the strong positive correlation would have been noticed.

@@

Table 1 summarises the model fits. It can be confusing the interpret the model estimates, and at first glance it appears the overlaid lines (-0.660) are substantially more powerful than the scatter plot (-4.403), but this is misleading, and  interpreting the the full model involves combining the estimates for the different interactions with correlation and sign. Taking correlation into account, the scatter plot shows superior performance. It needs to be noted that we only tested correlations at  0.3, 0.5, 0.7 and 0.9, so the results below an absolute value of 0.3 are unreliable and may be artifacts of regression smoothing methods. The correlation:sign interaction shows that the scatter plot has close to symmetric performance for positive (correlation:sign + correlation) and negative (-correlation) correlation while the overlaid line graph is substantially weaker for negative values of correlation. Surprisingly, smoothing out the excess 'noise' from a time series had a negative effect on scatter plots, despite a minimal visual impact of smoothed (Figure~\ref{smoothing}). It did have a significant positive impact on the power of overlaid line graphs. Increasing the sample size, $t$, improved the power of the scatter plot but it was not statistically significant for the overlaid line graph.
Figure~\ref{powermodel} shows the fitted unrestricted model, with power predictions (y axis) plotted against correlation (x axis) for each of the four sample sizes tested. The thicker lines represent the fixed effect fit, or the average performance across all tested individuals whilst the thinner lines represent the random effects fit, the variation in power that can be attributed to an individual's visual skill. Previous studies [15] show that this random visual skill is not impacted by demographics such as age and gender; whilst education increases the power of a graphic display by only approximately one percent, not a practically significant amount. We did not include these characteristics of the individual in the modelling due to this lack of significance. Figure~\ref{powermodel} especially highlights the weak performance of the overlaid lines and the benefits of smoothing.
